A comparison of design-based and model-based approaches for spatial data.

### 2021-08-13 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
minor updates

  \begin{itemize}
  \tightlist
  \item
    normal: mean 0, variance 2
  \item
    lognormal: log scale mean 0, log scale variance 2
  \item
    lognormalbig: log scale mean 0, log scale variance 4
  \end{itemize}
\item
  sample sizes: \(n = 10, 50, 150; N = 900\)
\item
  layout: gridded vs random population locations
\end{itemize}

@@

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}
\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-2.pdf}

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}

@@

@@

\begin{itemize}
\tightlist
\item
  In terms of rmspe, a model-based analysis on a GRTS design yields an
  rmspe similar to or lower than a design-based analysis on a GRTS
  design, as long as the response variable is not ``too skewed.''
\item
  If the response variable is very skewed, then neither analysis is
  appropriate, but, the design-based analysis is quicker.
\item
  a spatially balanced GRTS sample outperforms IRS in nearly all
  dependent error settings, as expected.
\item
  methods that use spatial correlation generally perform better on
  random location points than they do on gridded points. This makes some
  intuitive sense because (1) on average, the minimum distance between
  an unobserved point and its nearest observed neighbor should be lower
  for random points and (2) the span of the study area is maximized for
  a grid based on the way that we set up the simulations (with the
  random points being drawn as uniform random variables within the
  boundary of the grid).
\item
  comparison of Design-GRTS and Model-GRTS between two settings with
  different locations of points, but otherwise the same simulation
  parameters, should really be done on the same surface realization. One
  very strange realized response vector could drastically alter the
  results, especially on the exponentiated log data. In the same way
  that we compare the four approaches on the same realized data, we
  should also try to do the same with the locations, if they are of
  interest. (The realized mean won't be exactly the same but should be
  close).
\end{itemize}

\textbf{Sample Simulation}

For the following simulation results, we simulated 1040 different
gridded populations, each of size 900 (on the unit square) with sample
size 150. For the design-based approach, population units were selected
via GRTS, the Horvitz-Thompson estimator was used,and the local mean
variance was used. For the model-based approach (FPBK), population units
were selected via Independent Random Sampling (IRS) and the appropriate
prediction and prediction variance formulas were used.

The response was normally distributed with an exponential covariance
function with partial sill of 0.9, effective range of \(\sqrt{2}\), and
a nugget of 0.1. For model-based, we assumed the correct form of the
covariance function (exponential), but estimated the spatial parameters
with REML.

\hypertarget{software}{%
\subsection{Software}\label{software}}

>>>>

  \begin{itemize}
  \tightlist
  \item
    normal: mean 0, variance 2
  \item
    lognormal: log scale mean 0, log scale variance 2 (total variance
    47)
  \item
    lognormalbig: log scale mean 0, log scale variance 4 (total variance
    2,926)
  \end{itemize}
\item
  sample sizes: \(n = 10, 50, 150; N = 900\)
\item
  layout: gridded vs random population locations
\end{itemize}

@@

There were 2000 trials for each simulation. The original response
(before exponentiating if applicable) for the dependent error cases was
normally distributed with an exponential covariance function with
partial sill of 0.9, effective range of \(\sqrt{2}\), and a nugget of
0.1. For the independent error cases, the partial sill was 0 and the
nugget was 1.

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-4-1.pdf}

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}
\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-5-2.pdf}

@@

@@

\includegraphics{SpatialDVM_Manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}

@@

\begin{itemize}
\tightlist
\item
  In terms of rmspe, a model-based analysis on a GRTS design yields an
  rmspe similar to or lower than a design-based analysis on a GRTS
  design, as long as the response variable is not ``too skewed.''
\item
  If the response variable is very skewed, then neither analysis is
  appropriate, but, the design-based analysis is better.
\item
  a spatially balanced GRTS sample outperforms IRS in nearly all
  dependent error settings, as expected.
\item
  methods that use spatial correlation generally perform better on
  random location points than they do on gridded points. This makes some
  intuitive sense because (1) on average, the minimum distance between
  an unobserved point and its nearest observed neighbor should be lower
  for random points and (2) the span of the study area is maximized for
  a grid based on the way that we set up the simulations (with the
  random points being drawn as uniform random variables within the
  boundary of the grid).
\item
  comparison of Design-GRTS and Model-GRTS between two settings with
  different locations of points, but otherwise the same simulation
  parameters, should really be done on the same surface realization. One
  very strange realized response vector could drastically alter the
  results, especially on the exponentiated log data. In the same way
  that we compare the four approaches on the same realized data, we
  should also try to do the same with the locations, if they are of
  interest. (The realized mean won't be exactly the same but should be
  close).
\end{itemize}
